Likelihoods for a general class of ARGs under the SMC

### 2025-04-23 paper.tex
Changed my mind, backwards-in-time

% coalescent introduction: backwards-in-time
% left-to-right: Wiuf-Hein (difficult)
% SMC: although easy to reason about left-to-right sense
%
The coalescent process describes the ancestry of a set of sampled genomes.
The original algorithm
to simulate the coalescent with
recombination was formulated backwards in time \revpoint{2}{24}
by Hudson \citeyearpar{hudson_properties_1983}.
Wiuf and Hein later described
a tree-by-tree method to simulate the same stochastic process
\citep{wiuf_recombination_1999}.
This along-the-genome approach is considerably more complex as the next
tree to be simulated depends on all previous trees.
This idea however provided the basis of the
SMC \citep{mcvean_approximating_2005}, restricting the set of all state
space transitions possible under the Hudson coalescent to
obtain a process that is Markovian both backwards-in-time,
as well as along-the-genome. \revpoint{2}{25}
In its backwards-in-time formulation, the SMC only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineages with \emph{overlapping}
genomic intervals may coalesce. The SMC' is a refinement
of the SMC which also allows lineages with abutting genomic intervals \revpoint{2}{26}
to coalesce~\citep{marjoram_fast_2006}.
Although \citet{mcvean_approximating_2005} verbally described the SMC
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,
rasmussen_genome-wide_2014}.

@@

We now define the process more formally, by reintroducing
the notation of \citet{mcvean_approximating_2005} required to describe
the likelihood computation. This section
does not deviate from the original description of the SMC apart from
how coalescable pairs are counted. This subtle nuance
greatly simplifies the ability to keep track of coalescable pairs going
backwards in time.
At any point in time, the state of the process is
the set of labelled lineages extant at time $t$,
$L(t) = \{X_j(t)\}$.
A lineage $X_j(t) = [x_{j}, y_{j})$ is labelled by
an integer $j$ and defines the
half-open genomic interval of its ancestral material.
% Ea
% where each lineage $X_j(t)$ is labeled by an integer $j$
% and represented by a
% ancestral material.
% Any lineage can thus be summarized as
% for some interval endpoints $x_{j} < y_{j}$.
Because we look backwards in time, $t=0$ is today, and
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

@@

% Finally, by describing the SMC
% backwards in time, some intervals contained within the half-open
% interval representing a lineage might have fully coalesced before all the genetic
% material carried by that lineage has. As noted by \citet{mcvean_approximating_2005},
% such intervals can be observed as the stalks on the marginal genealogies (e.g., marginal trees
% for $[25, 81)$ Fig. \ref{fig:smc-unary}). Recombination events occurring within these smaller
% intervals however do not affect the marginal genealogies.

@@

% [JK: this needs significant work]
% Here, we have introduced a scalable algorithm to compute likelihoods
% under the SMC for a general class of ARGs. By formulating the SMC backwards in time
% we were able to associate a single likelihood value with each edge in the ARG.
% The compatibility of the algorithm with the \texttt{tskit} library allows us to
% exploit its associated efficiencies both for the current work and its future
% applications. This work will contribute significantly
% to ARG-based analysis becoming a standard part of the population
% and statistical genetics toolkit.
% In particular, we see two main applications for the work presented here.
% Firstly, the ability to associate inferred ARGs with a likelihood given a demographic
% model and an associated set of parameters, will help bridge the gap between the tremendous recent
% advances in ARG inference and the current state of the art statistical inference of
% past demography.

@@

% Since we look backwards in time, $t=0$ is today, and so

>>>>

% coalescent introduction: backwards-in-time
% left-to-right: Wiuf-Hein (difficult)
% SMC: although easy to reason about left-to-right sense
%
The coalescent process describes the ancestry of a set of sampled genomes.
The original algorithm
to simulate the coalescent with
recombination was formulated backwards-in-time \revpoint{2}{24}
by Hudson \citeyearpar{hudson_properties_1983}.
Wiuf and Hein later described
a tree-by-tree method to simulate the same stochastic process
\citep{wiuf_recombination_1999}.
This along-the-genome approach is considerably more complex as the next
tree to be simulated depends on all previous trees.
This idea however provided the basis of the
SMC \citep{mcvean_approximating_2005}, restricting the set of all state
space transitions possible under the Hudson coalescent to
obtain a process that is Markovian both backwards-in-time,
as well as along-the-genome. \revpoint{2}{25}
In its backwards-in-time formulation, the SMC only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineages with \emph{overlapping}
genomic intervals may coalesce. The SMC' is a refinement
of the SMC which also allows lineages with abutting genomic intervals \revpoint{2}{26}
to coalesce~\citep{marjoram_fast_2006}.
Although \citet{mcvean_approximating_2005} verbally described the SMC
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,
rasmussen_genome-wide_2014}.

@@

We now define the process more formally, by reintroducing
the notation of \citet{mcvean_approximating_2005} required to describe
the likelihood computation. This section
does not deviate from the original description of the SMC apart from
how coalescable pairs are counted. This subtle nuance
greatly simplifies the ability to keep track of coalescable pairs going
backwards-in-time.
At any point in time, the state of the process is
the set of labelled lineages extant at time $t$,
$L(t) = \{X_j(t)\}$.
A lineage $X_j(t) = [x_{j}, y_{j})$ is labelled by
an integer $j$ and defines the
half-open genomic interval of its ancestral material.
% Ea
% where each lineage $X_j(t)$ is labeled by an integer $j$
% and represented by a
% ancestral material.
% Any lineage can thus be summarized as
% for some interval endpoints $x_{j} < y_{j}$.
Because we look backwards-in-time, $t=0$ is today, and
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

@@

% Finally, by describing the SMC
% backwards-in-time, some intervals contained within the half-open
% interval representing a lineage might have fully coalesced before all the genetic
% material carried by that lineage has. As noted by \citet{mcvean_approximating_2005},
% such intervals can be observed as the stalks on the marginal genealogies (e.g., marginal trees
% for $[25, 81)$ Fig. \ref{fig:smc-unary}). Recombination events occurring within these smaller
% intervals however do not affect the marginal genealogies.

@@

% [JK: this needs significant work]
% Here, we have introduced a scalable algorithm to compute likelihoods
% under the SMC for a general class of ARGs. By formulating the SMC backwards-in-time
% we were able to associate a single likelihood value with each edge in the ARG.
% The compatibility of the algorithm with the \texttt{tskit} library allows us to
% exploit its associated efficiencies both for the current work and its future
% applications. This work will contribute significantly
% to ARG-based analysis becoming a standard part of the population
% and statistical genetics toolkit.
% In particular, we see two main applications for the work presented here.
% Firstly, the ability to associate inferred ARGs with a likelihood given a demographic
% model and an associated set of parameters, will help bridge the gap between the tremendous recent
% advances in ARG inference and the current state of the art statistical inference of
% past demography.

@@

% Since we look backwards-in-time, $t=0$ is today, and so
